---
title: "Factorial ANOVA"
author: "Danielle van Os"
date: "21/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stats) # For linear models
library(MASS) # For random data generation with exact means 
library(knitr) # for nice tables

rm(list = ls())
```



# Two-way/Factorial ANOVA 

The model for a two-way ANOVA with interactions includes:

* Main effect of variable 1   
* Main effect of variable 2  
* Interactions between each level of variable 1 and level 2  
* error 

## Example equations

### Example 1 - 2x3 with interaction
Variable 1 has two levels, variable 2 has three levels, with an interaction term:

$$ y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_1X_2 $$
Where:

* $\beta_0$ is the grand mean, or the mean for the first level of all factors  
* $X_1$ is the score on variable 1 
* $X_2$ is the score on variable 2 
* $\beta$ represents the slope or beta value 

NOTE TO SELF: CHANGE THE CODE BELOW SO THERE IS AN OPTION TO HAVE INTERACTIONS OR NO INTERACTION    

What is an interaction? Broadly, it suggests that the effect of one predictor variable on the outcome variable is dependent on the other predictor. 

```{r}
# THE ISSUE WITH THIS WAY IS THAT VAR1 AND VAR2 ARE ADDED TOGETHER TO CREATE A SINGLE VALUE (RANDOM, NO PARTICULAR REASON, I JUST NEEDED A SINGLE VALUE) BUT ADDING THEM MEANS THEY'RE SINGULAR SO COME UP WITH AN ALTERNATIVE WAY TO MODEL BELOW 
# Parameters
total_sample_size <- 600

var1_nvar <- 2
var1_g1_mean <- 10
var1_g2_mean <- 12
var1_g1_sd <- 0.3
var1_g2_sd <- 0.3
 
var2_nvar <- 3
var2_g1_mean <- 3
var2_g2_mean <- 6
var2_g3_mean <- 8
var2_g1_sd <- 0.3
var2_g2_sd <- 0.3
var2_g3_sd <- 0.3 

 

# Data generation
var1_g1 <- mvrnorm(n = round(total_sample_size/var1_nvar), mu = var1_g1_mean, Sigma = var1_g1_sd, empirical = TRUE)
var1_g2 <- mvrnorm(n = round(total_sample_size/var1_nvar), mu = var1_g2_mean, Sigma = var1_g2_sd, empirical = TRUE)
var1 <- c(var1_g1, var1_g2)

var2_g1 <- mvrnorm(n = round(total_sample_size/var2_nvar), mu = var2_g1_mean, Sigma = var2_g1_sd, empirical = TRUE)
var2_g2 <- mvrnorm(n = round(total_sample_size/var2_nvar), mu = var2_g2_mean, Sigma = var2_g2_sd, empirical = TRUE)
var2_g3 <- mvrnorm(n = round(total_sample_size/var2_nvar), mu = var2_g3_mean, Sigma = var2_g3_sd, empirical = TRUE)
var2 <- c(var2_g1, var2_g2, var2_g3)

# Data frame
df <- tibble(var1_label = rep(c("group_1", "group_2"), each = total_sample_size/var1_nvar),
             var2_label = rep(c("group_1", "group_2", "group_3"), each = total_sample_size/var2_nvar),
             
             y = c(var1 + var2),
             
             var1_group_n = as_factor(rep(c(1, 2), each = total_sample_size/var1_nvar)),
             var2_group_n = as_factor(rep(c(1, 2, 3), each = total_sample_size/var2_nvar)),
             
             var1_g2 = ifelse(var1_label == "group_2", 1, 0),
             var2_g2 = ifelse(var2_label == "group_2", 1, 0),
             var2_g3 = ifelse(var2_label == "group_3", 1, 0)
             )

# Linear model and anova function 
summary(aov(y ~ var1_label + var2_label, data = df))

fact_anova1 <- summary(lm(formula = y ~ 1 + var1_g2 + var2_g2 + var2_g3 + var1_g2*var2_g2 + var1_g2*var2_g3, data = df))
fact_anova2 <- summary(lm(formula = y ~ var1_label * var2_label, data = df))

fact_anova1
fact_anova2
# Output
anova

knitr::kable(coefficients(anova), digits=3)
```

## 2x2 ANOVA 

```{r}

# Set coefficients
alpha = 10
beta1 = .3
beta2 = -.5
beta3 = -1.1

# Generate 200 trials
A = c(rep(c(0), 100), rep(c(1), 100)) # '0' 100 times, '1' 100 times
B = rep(c(rep(c(0), 50), rep(c(1), 50)), 2) # '0'x50, '1'x50, '0'x50, '1'x50
e = rnorm(200, 0, sd=1) # Random noise, with standard deviation of 1

# Generate your data using the regression equation
y = alpha + beta1*A + beta2*B + beta3*A*B + e

# Join the variables in a data frame
data = data.frame(cbind(A, B, y))

# Fit an ANOVA
model = aov(y ~ A*B, data=data)
summary(model)
```

## 3x2 ANOVA
```{r}
# Set coefficients
alpha = 10
beta1 = .3
beta2 = -.5
beta3 = -1.1
beta4 = 0.5

## Generate data 
# 3x2 balanced design
A = c(0,0,0,0,1,1,1,1)
B = c(0,0,1,1,0,0,1,1)
C = c(0,1,0,1,0,1,0,1)

# Repeat 50 times, for N = 200
A = rep(A, 50)
B = rep(B, 50)
C = rep(C, 50)
e = rnorm(400, 0, sd=1) # Random noise, with standard deviation of 1 

# Generate your data using the regression equation
# NOTE TO SELF: should I include two-way interactions here??? 
y = alpha + beta1*A + beta2*B + beta3*C + beta4*A*B*C + e

# Join the variables in a data frame
data = data.frame(cbind(A, B, C, y))

# Fit an ANOVA
model = aov(y ~ A*B*C, data=data)
summary(model)

```





```{r}

y = 1*alpha + var1_g2 + var2_g2 + var2_g3 + var1_g2*var2_g2 + var1_g2*var2_g3

# Join the variables in a data frame
data = tibble(A, 
              B, 
              y,
              A_g2 = ifelse(A == 1, 1, 0),
              B_g2 = ifelse(B == 1, 1, 0),
              B_g3 = ifelse(B == 2, 1, 0))

# Fit an ANOVA
model = aov(y ~ A*B, data=data)
summary(model)

# Linear model equivalent 
"A"
A
"B"
B
```

In R it is written as follows: 
```{r class.source = 'fold-show', eval = FALSE}

lm(outcome_var ~ 1 + # the 1 is included to indicate $1*B0/intercept
   var1_level2 + var2_level2 + var2_level3 + # main effects 
     var1_level2*var2_level2 + var1_level2*var2_level3) # interaction terms 
  
```


Note the first group on both variables is not included.

### - 2x3 without interaction


### - 4x5 with interaction

It quickly gets difficult to model each 











