---
title: "Logistic Regression"
author: "Danielle van Os"
date: "02/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Remove previous environment
rm(list = ls())

# Load libraries
library(tidyverse)
library(MASS) # mvrnorm


```

To use the general linear model for binary outcome variables (e.g., yes/no) you need a logistic link, which gives a logistic regression model. 

Recall the regression equation: 

$$ f(x) = \alpha + \beta_1x + ... + \beta_px$$

And the logistic function: 

$$\frac {1}{1 + e^{-x}}$$
Where:  
* 1 in the numerator is the curve's maximum value

In the case of logistic regression, the regression equation is substituted into the 'x' value in the logistic function: 

$$f(x) = \frac {1}{1 + e^{-(\alpha + \beta_1x + ... + \beta_px)}}$$


## Standard Logistic Regression

```{r}

# Set parameters 
sample_size <- 300
intercept <- 10
slope <- 40
sd <- 0.1
mean <- 0

# Logistic function
logistic <- function(x){
  return(1 / (1 + exp(-x)))
}


# Simulate data 
# x value
x <- mvrnorm(n = sample_size, mu = 0, Sigma = sd, empirical = TRUE) 

# Regression equation to sub into logistic function
regression_eq <- intercept + slope*x + mvrnorm(n = sample_size, mu = 0, Sigma = sd, empirical = TRUE) 

# y value 
y <- logistic(regression_eq)

# Creating the tibble 
df <- tibble(x = x, y = y)

# Running the model 
lm <- summary(lm(formula = y ~ x, data = df))
lm

```

```{r}

ggplot(data = df, mapping = aes(x = x, y = y)) + 
  geom_point()

```

## Logistic Regression with Categorical Predictor Variables 

Resources:  
* https://stats.stackexchange.com/questions/3386/how-to-deal-with-non-binary-categorical-variables-in-logistic-regression-spss  
* https://stats.stackexchange.com/questions/225058/logistic-regression-with-only-categorical-predictors  

```{r}

```

## Multinomial Logistic Regression 

```{r}

```

