---
title: "T-Test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stats) # For linear models
library(MASS) # For random data generation with exact means 
library(knitr) # for nice tables

rm(list = ls())
```



## t-test - 1 variable 2 groups

```{r}
# Parameters
group_size <- 200 
mean_g1 <- 10
mean_g2 <- 12
sd_g1 <- 0.3
sd_g2 <- 0.3

# Data generation
group_1 <- mvrnorm(n = group_size, mu = mean_g1, Sigma = sd_g1, empirical = TRUE)
group_2 <- mvrnorm(n = group_size, mu = mean_g2, Sigma = sd_g2, empirical = TRUE)
data <- c(group_1, group_2)

# Data frame
df <- tibble(group = rep(c(0, 1), each = group_size),
             y = data)

# Linear model and t.test function 
t.test(formula = y ~ group, data = df)
t_test <- summary(lm(formula = y ~ 1 + group, data = df)) # The '1' is a way of writing 1*B0 or 1*intercept, which in this case is 1*10. 

# Output
t_test

knitr::kable(coefficients(t_test), digits=3)

```
The x estimate indicates the amount that the second group mean differs from the first. If it is negative it means that the mean of the second group is less than that of the first group, whereas if it is positive, the mean of the second group is greater than that of hte first group. 

### Plot T-Test 

NOTE TO SELF: Eventually animate the transition from a standard graph to the density plots 
NOTE 2: add intercept horizontal line as well 

```{r graph t-test}

# Step 1 - standard graph
ggplot(data = df, mapping = aes(x =group, y = y)) + 
  geom_point() + 
  stat_smooth(method = "lm")

# Step 2 - change axes 
ggplot(data = df, mapping = aes(x = y, y =group)) + 
  geom_point() + 
  stat_smooth(method = "lm")

# Step 3 - remove regression line 
ggplot(data = df, mapping = aes(x = y, y =group)) + 
  geom_point() 

# Step 4 - change colours 
ggplot(data = df, mapping = aes(x = y, y =group, color =group)) + 
  geom_point(alpha = 0.5) + 
  theme(legend.position = "none")

# Step 5 - move dots from second graph to the x axis 
ggplot(data = df, mapping = aes(x = y, y = 0, color =group)) + 
  geom_point(alpha = 0.5) + 
  ylim(0, 1) + 
  theme(legend.position = "none")

# Step 6 - turn into density plots for easier viewing (how t-tests are usually displayed)
ggplot(data = df, aes(x = y, group =group, fill =group)) + 
  geom_density(alpha = 0.5) + 
  theme(legend.position = "none")

```
